---
title: "Model of translational coupling"
output:
  pdf_document:
    keep_tex: yes
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## A model of translational coupling

To model the effects of translational coupling on protein production, we first assume a polycistronic transcript with three genes _a_, _b_, and _c_. For the effective initiation rate of $a_\text{in}$ we define the following relationship
$$a_\text{in} = \min\left\{i_a, \tau_a\right\}$$
where $i_a$ is the aggregate initiation rate of _a_ and $\tau_a$ is the translation elongation rate of gene _a_. We assume that if the initiation rate ever exceeds the elongation rate, ribosomes would quickly back up on the transcript and make elongation the rate-limiting step of translation. Thus in our model, the elongation rate can never be exceed by the aggregate translation initiation rate. For gene _a_, the aggregate initiation rate is simply the _de novo_ initiation rate because there are no genes upstream of _a_.

The rate at which ribosomes complete translation of gene $a$, or protein production rate of A, is defined as

$$ \dot{A} = a_\text{in} $$
at steady state. In this context, the steady state assumption means that all three protein products are being formed continuously.

For the effective translation initiation rate $b_\text{in}$ of gene _b_, we similarly write
$$b_\text{in} = \min\left\{i_b, \tau_b\right\}$$

where $i_b$ is the aggregate translation initiation rate due to upstream-dependent reinitiation and \textit{de novo} initiation, and $\tau_b$ is the translation elongation rate of _b_. We define the aggregate translation initiation rate
$$i_b = b_\text{reinit} + b_\text{de novo}$$
where $b_\text{reinit}$ is the rate of upstream translating ribosomes reinitiating on gene \textit{b} and $b_\text{de novo}$ is the rate of ribosomes initiating _de novo_ on gene _b_.

Lastly we define reinitiation and _de novo_ initiation rates on gene _b_ as follows

$$b_\text{reinit} = q_b\dot{A}$$
$$b_\text{de novo} = z_b\dot{A} + w_b$$
where $\dot{A}$ is the rate of ribosomes flowing from upstream translation of gene $a$, and $q_b$ represents the proportion of that ribosome flow reinitiating on gene _b_. We assume that the rate of _de novo_ initation depends, in part, on upstream ribosomes relaxing secondary structure around the ribosome binding site of gene _b_. Thus, the rate of _de novo_ initiation is given by the upstream ribosome flow $\dot{A}$ scaled by some constant $z_b$, and by a constant rate $w_b$ that does not depend on upstream ribosome flow.

We can simplify the effective initiation rate to
$$b_\text{in} = \min\left\{\dot{A}(q_b + z_b) + w_b, \tau_b\right\}$$

We simplify the effective initiation rate further by defining a coupling constant 
$$
y_b = q_b + z_b
$$
which accounts for both the effects of facilitated binding and ribosome reinitiation. The final effective initiation rate of _b_ is defined as

$$
b_\text{in} = \min\left\{y_b\dot{A} + w_b, \tau_b\right\}
$$

Similar to the protein production rate of gene _a_, we define the protein production of gene _b_ 

$$ \dot{B} = b_\text{in} $$

at steady state.

The effective initiation rate of gene _c_ is similar to that of gene _b_

$$
c_\text{in} = \min\left\{y_c\dot{B} + w_c, \tau_c\right\}
$$
$$
y_c = q_c + z_c
$$

where $\dot{B}$ is the rate of ribosomes flowing from the end of upstream gene $b$, and $q_c$ represents the proportion of that ribosome flow reinitiating on gene _c_. The rate of _de novo_ initiation dependent on upstream ribosome flow is given by $z_c$, and $w_c$ is the _de novo_ initation rate independent of upstream ribosome flow. Again, $y_c$ is a coupling constant that incorporates the effects of both facilitated binding and ribosome reinitiation.

### Results and figures

I have made plots similar to Fig 7 in our paper over a range of $y$ and $w$ values at steady state. We assume that $y_c = y_b$ and $w_b = w_c$. We also assume that $\tau_a = \tau_c$, and we vary $\tau_b$.

```{r model_fig, echo=FALSE, warning=FALSE, message=FALSE, fig.width=10, fig.height=6}
library(dplyr)
library(tidyr)
library(ggplot2)
library(cowplot)

k <- 3
tau_1 <- 3.5
tau_3 <- 3.5

a_in <- function() min(k, tau_1)
a_dot <- function() {a_in()/a_in()}
b_in <- function(tau_2, y, w) min(a_dot()*(y) + w, tau_2)
b_dot <- function(tau_2, y, w) {b_in(tau_2, y, w)/b_in(8, y, w)}
c_in <- function(tau_2, y, w) min(b_dot(tau_2, y, w)*(y) + w, tau_3)
c_dot <- function(tau_2, y, w) {c_in(tau_2, y, w)/c_in(8, y, w)}

taus <- seq(0, 8, 0.01)


make_df <- function(y, w) {
  data.frame(rate_B = taus,
             A = a_dot(),
             B = sapply(taus, b_dot, y, w),
             C = sapply(taus, c_dot, y, w),
             y = y,
             w = w)
}

y_vals <- list(0.1, 0.7, 1.2)
w_vals <- list(0.2, 0.4)

df <- purrr::cross_d(list(y_vals = y_vals, w_vals = w_vals)) %>% 
  mutate(data = purrr::map2(y_vals, w_vals, make_df)) %>% unnest()
df <- gather(df, protein, rate, A:C)

prepender_w <- function(string, prefix = "initiation: ") paste0(prefix, string)
prepender_y <- function(string, prefix = "coupling: ") paste0(prefix, string)

plot <- ggplot(df, aes(x = rate_B, y = rate, color = protein)) +
  geom_line() +
  facet_grid(w_vals~y_vals, labeller = labeller(w_vals = as_labeller(prepender_w), y_vals = as_labeller(prepender_y))) +
  panel_border() +
  xlab("translation elongation rate of protein B (1/s)") +
  ylab("relative rate of protein production") +
  xlim(0, 1.6)

plot

save_plot('../figures/model_plot.pdf', plot, base_aspect_ratio = 1.7)
```
